### 2021-10-06 submission/response_to_reviewers.md

**I read the manuscript  "Removal of rare amplicon sequence variants from 16S rRNA gene sequence surveys biases the interpretation of community structure data" from Dr. Schloss with great interest. There are a diversity of opinions about whether ASV or OTU tables should or should not be filtered or quality controlled prior to subsequent statistical investigation and an even greater diversity about the specific approaches that should be applied. This may hold important implications for whether and which patterns are observed in the analysis of microbiome data, yet there has been surprisingly little work to measure what the effects of these decisions actually are. Dr. Schloss' effort to evaluate the effect of these decisions in a data-driven way thus holds broad implications for the entire field of microbiome data analysis. In particular, Dr. Schloss analyzed a variety of publicly available datasets to determine how the removal of rare ASVs or OTUs affects the patterns present in microbiome data. This extensive analysis seeks to clarify whether the common approach of removing rare variants - which are often presumed to represent spurious sequences (i.e., artifacts) - from an analysis is meritorious. Dr. Schloss' analysis points towards two big conclusions: (1) rare variants add information to microbiome studies and should be retained and (2) rarefaction techniques should be used to control sequencing depth and thus the number of artifacts present in an analysis.**

**While these are important questions to address, I do not walk away from the manuscript - in its current form - convinced by its conclusions, especially given some of the strong language that is invoked. In particular, I think the manuscript could make a better case by improving its clarity in some areas, especially with respect to methodology and rationale, and by revising or adding analyses in other places to solidify conclusions. Moreover, some of the conclusions may require hedging given caveats that are not currently considered in the present analysis. Given that this manuscript upends long-standing practice and expectations, it seems especially important that it fill in these gaps.**

**Below, I itemize my criticisms with the manuscript. I hope Dr. Schloss sees these as constructive comments designed to help him improve this manuscript, which I ultimately think will play an important role in the advancement of microbiome data analytics.**

* Grateful that the reviewer sees value in this work and the questions I have raised.

* A significant challenge of this type of work is that we do not know the *true* answer for any of these communities. Any analysis that has been done regarding thresholds for removing sequences has been based on mock communities which are very much not realistic

* Point out that aside from those problematic mock community analyses there has been little to no work demonstrating that we *should* be removing singletons. That decision was made by fiat rather than based on data.

**1. It is currently challenging to evaluate the work because the methodological details used to conduct it are frequently sparsely described. The materials and methods are in effect integrated into the results section. While I do not have a problem with this per se, it seems to have come at the cost of not providing specific details about various analytical steps that were implemented as part of the analysis as well as the rationale for implementing these techniques. This ranges from relatively straightforward aspects of methods, such as the reason for why these specific data sets were selected (from among the others that fit the same criterion of targeting the V4 locus with the same sequencing technology), to whether multiple tests correction was applied to statistical tests. In other places, the methodology is confusing as described, especially regarding the various null model sampling approaches (for example, see my specific comment about line 200 below). While I appreciate that Dr. Schloss has made all code available to readers, I think it's important to ensure that an understanding of the approach by a reader does not depend upon their ability to read code. I therefore recommend that the technical approach be more explicitly described in the manuscript.**

* More justification for why studies were selected
* Have removed description of the methods from the Results section
* Statistical tests
* More detailed model descriptions

**2. I'm not sure that the manuscript provides evidence that rare sequences should be retained in an investigation, but rather documents the sensitivity of commonly applied analytical approaches to the decisions made about whether and how rare sequences should be incorporated. The former statement requires demonstrating that the inclusion of rare variants improves analytical accuracy. While the present analysis suggests that including rare variants impacts alpha- and beta-diversity metrics as well as the sensitivity of tests that utilize these metrics (according to the null model approach), it is not evident if the filtered or unfiltered data better reflects the underlying ecology. As a brief aside, Figure 5 offers the clearest insight into the potential impact of filtering on analytical accuracy, but it is not clear from the methodology how treatment groups were established and thus it is hard to follow Dr. Schloss' interpretation here.**

* Problem with knowing *truth*
* All models/simulations are worthless
* Figures 4 and 5 show that removing rare sequences causes problems for statistical power (Figure 4) and false detection of differences (Figure 5B)

**In general, much of the interpretation of the results appears predicated on the idea that the rare variants represent real taxa. However, there is only limited evidence in support of this view. In the section "The nature of singletons", Dr. Schloss finds that singletons are more prevalent in more shallowly sequenced samples and that singletons in a sample are often also present in another sample. Dr. Schloss concludes that these singletons are thus likely real variants and the rest of the investigation seems somewhat predicated upon this interpretation. However, I am left with several questions:**

**(A) Are the correlations meaningfully representative of a trend across all samples, or are they driven by a small number of outlier samples? Providing support of these trends by plotting the scatterplots for each study would help assuage this concern.**

* Used Spearman correlation coefficients, which are robust to the effects of outliers
* Create a new Figure S2 that shows the requested plots for the data in Fig 1AC

**(B) Despite these observations, couldn't these rare taxa include artifacts nonetheless and, if so, do real rare variants and artifacts contribute uniform information in all subsequent analyses? An effort to directly measure the impact of spurious sequences on analytical accuracy would help offset this concern.**

* Lengthy discussion of the fact that there are likely spurious sequences in the rare pool
* Value of rarefaction to control for the higher likelihood of crummy sequences with greater sequencing depth
* Make comparisons based on relative differences in diversity rather than absolute quantities

**(C) Since the processes underlying the generation of artifacts are not explicitly described, the reader is left unsure about the expectation that the formation of singleton artificats is driven by sequencing depth.**

* Hopefully the revised text will clarify these issues.

**3. I'm not sure that the description of variant filtering accurately reflects common practice. For example, while many groups appear to apply count thresholds to determine which taxa should be retained for analysis, these thresholds are typically applied across all samples and are often cross referenced with prevalence filtering. As a result, taxa that are singletons in one sample but present in others are typically included in an analysis despite otherwise being below the count threshold. In Dr. Schloss' study, however, such taxa are considered candidates for filtration, which raises questions about the applicability of the results. Moreover, frequent practice thresholds on relative abundance rather than direct counts to offset the potential for sequencing depths to drive singleton formation effects. Dr. Schloss may consider discussing or even experimentally measuring the effect of these alternative strategies on the outcomes of his analyses.**

* As outlined in the Introduction, sample-by-sample abundance filtering is the norm for current approaches to identifying ASVs
* Prevalence filtering...
* Relative abundance is effectively the same as abundance filtering. The relative abundance would indicate the average number of counts for the OTU across a large number of resamplings

**4. Related to the above, it is not clear if the results presented here are generalizable to other investigations. For example, it is possible that the use of the high-quality data sets included in this analysis or the specific bioinformatic techniques applied here results in a lower than typical rate of artifact formation compared to what many readers may encounter in their own analyses. For example, studies that do not apply library prep approaches that produce two overlapping V4 reads, but that rather amplify a longer amplicon target and thus have lower sequence accuracy, may have an increased rate of forming artifacts in their data. Likewise, ASVs in this study are sequences that tolerate a 2 bp difference, which may differ from the set of ASVs produced using another approach. Since the study never directly measures the cost of including artifactual sequences (2B above), it is not clear if the recommendations made here are meaningful for investigations that involve data subject to a greater rate of artifact formation.**

* I do not believe that authors have an excuse to generate data with low quality data. As outlined in the Discussion, this is part of the problem with the field currently.

**5. Conclusions around rarefaction need more justification or support. There are several strongly worded statements in the manuscript about the use of rarefaction techniques, but there appears to be little direct assessment of the impact of rarefaction on study outcomes. Additionally, there is no comparison with alternative rarefaction-free normalization techniques, including simply filtering based on relative abundance. Given the on-going debate in the literature about the value of rarefaction, it seems important that the manuscript solidify this point if the language is to be included as is.**

* Even Holmes's study says to rarefy alpha and beta diversity metrics
* Add citations to rarefaction yes papers
* Analysis of rarefaction is beyond the scope of this paper

**6. The rationale underlying some of the analyses, especially with respect to the null models, needs support to help readers connect the results with the conclusions. For example, for the section on line 122 ("The impact of removing rare ASVs on the information represented in each sample.") it seems self-evident that removing rare taxa will affect the estimate of alpha-diversity, especially when considering the well documented rank abundance distribution and long-tail of low abundance taxa that typically comprises microbial communities. Dr. Schloss may get more mileage out of this section by clarifying why this analysis is noteworthy, or by focusing of metrics that are less directly sensitive to the number of taxa (perhaps evenness).**

* Sobs is definitely obvious
* Bigger point in Figure 3 is that the coefficient of variation between samples goes up with removal of rare sequences reducing power to detect differences in communities
* Will also caculate Shannon evenness (keep in mind that this is dependent on richness)

**Additionally, I'm not sure that the treatment effect simulation analysis is as meaningful as indicated by the text (though again, I'm not sure I follow the approach properly). If the ASV sampling procedure to determine which ASVs are increased in abundance is simply selecting from a list of ASVs, then the long-tail of rare ASVs will tend to be selected versus those that tend to abundant. Consequently, when the procedure amplifies the abundance of these (and only these) selected ASVs, again which are biased towards rare taxa, then it will be rare taxa driving the differences between simulated treatments. Abundance-based filtering of said taxa will obviously have an effect on the analysis. Is this akin to what happens in microbial communities in nature? Well, maybe in some instances, but lots of studies find that it's the abundant taxa that differentiate treatments. Such studies may not need the singleton ASVs to observe the effect. Could those ASVs help clarify the effect? Perhaps. But inclusion of artifact sequences (which again are not directly tested here) could theoretically increase the error of the analysis in such cases at a greater rate. I would recommend considering a parallel analysis that simulates ASVs based on their abundance to help clarify whether the interpretation drawn from this section is relevant to communities that behave along the lines described above. Alternatively, hedge the conclusion by clarifying the above caveat, but that this analysis may yet reveal insights into how rare ASVs could reveal differences currently being missed in communities that do not substantially change with respect to the abundant fraction of community membership.**

* As the reviewer points out, we don't know. There will certainly be times where differences between communities are driven by rare populations and times where it splits by abundant populations. We really don't know. Those that are driven by abundant populations will be obviously different already so it's a moot point for that case
* High quality data reduces risk of artifacts and rarefaction allows us to control for inclusion of artifacts

**Minor comments:**

**Line 94: What is meant by "relatively small" when referring to the marine and soil data sets? Based on table 1, it seems to be the number of samples. But you can help readers by being more explicit here especially given that the surrounding context of discussion pertains to sequencing depth.**

* Yes number of samples. Clarify text

**Line 102: "In addition, with the exception of the samples from the marine and sediment datasets (Spearman correlation, P>0.05)..." Indicate parenthetically here that these samples were non-significant (as opposed to the immediate assumption that they are positively associated) and move the parenthetical pointer to spearman and the pvalue to later in the sentence.**

* Clarify text

**Line 148: "For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness  and diversity of ASVs and one would not expect to find treatment-based effects between the samples." It would be more compelling to demonstrate that this expectation was met in the resulting distributions. Can you offer graphical or statistical evidence in support of this?**

* That is shown in Figure 3 when rare sequences were not removed

**Line 175: How many samples are being selected for each simulated treatment in this analysis?  At first I thought all samples in a study were selected, but it looks like this isn't the case, as an undescribed number of "communities were randomly sampled from the null distribution..."**

* Clarify that the samples were sampled into one of two equally sized treatment groups

**Line 179: What p-value threshold is being applied to "detect a difference" and quantify the "fraction of significant tests"?**

* Clarify that it is P<0.05

**Line 200: It's a little challenging to determine what happened here: "First, I sampled sequences from the 200 null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs". For example, what is meant by "sequences" here? ASVs? How did you sample the sequences? Are you saying that you subsampled the data somehow? I recommend you clarify.**

* Clarify text

**Line 204: What is meant by "meaningfully vary"? Was this quantified in some manner?**

* Indicate a range? Seems clear in Fig 5A and S6A

**Line 215: "However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity". But Fig 5B seems to show that the rate only tends to increase when singletons are removed, such that the rate is flat as more abundant rare sequences are removed.**

* Yes it's flat, but the rate increased when any rare sequences were removed. That's the problem. The value is the lowest when all sequences were retained.

**Line 236: "The practice of removing rare sequences from samples seems to be a response to researchers prioritizing the number of reads and length of sequences over their quality." I'm not sure this is entirely accurate. Some of it stems from this, but it also appears to have emerged as a way to control for overclustering, which often results in spurious OTUs. This methodology may have been a holdover in the era of ASV analysis, where it may not be as critical.**

* I stand by this comment

**Line 250: "In practice, I suspect this range is actually larger since researchers may have opted against depositing samples with fewer reads into the SRA". Perhaps. But this would imply that such samples were not (nor would be) included in the analysis and not subject to the bias being discussed here.**

* Sentence removed

**Line 255: I think I follow your intention, but note that your results of finding that rare sequences are higher in shallow depth samples seems initially at odds with the subsequent part of the discussion that observes that "number of spurious sequences increases with sequencing depth". You can avoid some confusion by using this observation is the expectation and then discussing how your results differ from this expectation, reinforcing the interpretation that the rare sequences observed in your analysis are real variants. This may be your intention in the current narrative; I simply recommend a modest restructure of this logic to improve comprehension.**

* Clarify text

**Line 287: Typo. Change insure -> ensure**

* Corrected

### 2021-10-21 submission/manuscript.Rmd
Add running title
Show raw data that went into correlations

@@

```{r}

frac_lost <- read_tsv(here('data/process/sequence_loss_table_raw.tsv')) %>%
	filter(min_freq == 1) %>%
	group_by(dataset) %>%
	summarize(median=median(fraction_lost))

@@

cor_n <- read_tsv(here('data/process/sequence_loss_table_cor.tsv')) %>%
 	filter(min_freq == 1) %>%
	select(dataset, cor_n_estimate, cor_n_p.value) %>%
	mutate(sig = cor_n_p.value < 0.05)

@@

frac_coverage <- read_tsv(here('data/process/sequence_coverage_table_raw.tsv')) %>%
	filter(min_freq == 1) %>%
	group_by(dataset) %>%
	summarize(median=median(coverage))

@@

coverage_n <- read_tsv(here('data/process/sequence_coverage_table_cor.tsv')) %>%
	filter(min_freq == 1) %>%
	select(dataset, cor_n_estimate, cor_n_p.value) %>%
	mutate(sig = cor_n_p.value < 0.05)

@@

**The nature of singletons.** Removal of rare ASVs is commonly justified as a method of removing ASVs that are artifacts. If such ASVs are artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth. Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`). In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figure 1A). This showed that with additional sequencing, the probability of seeing singleton ASVs in multiple samples was greater than the probability of generating an artifact. This suggests that the singleton ASVs are not as likely to be artifacts as previously thought. Furthermore, if singleton ASVs were artifacts, then one would not expect to find them in other samples from the same dataset. In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences. At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset (Figure 1B). Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable. The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively (Figure 1C). The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples. Among the three datasets without a significant correlation (Spearman correlation, P>0.05), the marine and soil datasets had the fewest samples in our collection and the stream dataset already had a high level of coverage regardless of the number of sequences. Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes more non-artifact ASVs than those that are artifacts.

```{r}
ointra <- read_tsv(here("data/process/ointra_analysis.tsv"),
		col_types = cols(.default=col_double(),
					dataset=col_character(), method=col_character(), group=col_character())
	) %>%
	group_by(dataset, method, min_class) %>%
	summarize(mean_s_frac = 100*(1-mean(s_yes / s_no)),
		mean_h_frac = 100*(1-mean(h_yes / h_no))) %>%
	ungroup()

@@

**The impact of removing rare ASVs on the information represented in each sample.** Removing rare ASVs will reduce the richness of ASVs (i.e. the number of ASVs per sample) and increase the relative abundance of the remaining ASVs. To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample. The richness of ASVs in each sample decreased by between `r oxford_comma(pull(asv_s_two, mean_s_frac), 1)`% (`r oxford_comma(pull(asv_s_two, dataset))`) when removing those ASVs that only appeared once and by between `r oxford_comma(pull(asv_s_eleven, mean_s_frac), 1)`% (`r oxford_comma(pull(asv_s_eleven, dataset))`) when removing those that appeared ten or fewer times from each sample (Figure 2A). Similarly, the Shannon diversity decreased by between `r oxford_comma(pull(asv_h_two, mean_h_frac), 1)`% (`r oxford_comma(pull(asv_h_two, dataset))`) when removing ASVs that only appeared once and by between `r oxford_comma(pull(asv_h_eleven, mean_h_frac), 1)`% (`r oxford_comma(pull(asv_h_eleven, dataset))`) when removing ASVs that appeared ten or fewer times from each sample (Figure 2B). Next, I assigned the ASVs to OTUs, which were defined as a group of ASVs that were more than 97% similar to each other to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies. Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the richness of OTUs and the diversity decreased relative to the full community (Figure S2AB). In contrast to the richness and diversity measurements, the Kullback–Leibler divergence compares the relative abundance of specific ASVs or OTUs between representations of the community. I calculated the Kullback–Leibler divergence between the full communities and those where rare ASVs were removed. As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 2C and Figure S2C). The relative loss of information was generally smaller for OTUs than than it was for ASVs. Removing rare ASVs, regardless of abundance threshold, had profound impacts on the representation of the communities.

**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing, without replacement, the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed. This effectively made every community in a study a statistical sample of the study-wide composite community distribution. For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples. Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study. The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S3). The null distribution data were used in the remainder of the study to minimize the risk of bias.

```{r}
r_cov_alpha <- read_tsv(here("data/process/ralpha_analysis.tsv"),
		col_types = cols(.default=col_double(),
					dataset=col_character(), method=col_character())
	) %>% select(dataset, method, prune, starts_with("cv_")) %>%
	filter(prune == 2 | prune ==1) %>%
	pivot_wider(id_cols=c(dataset, method), names_from=prune, values_from=starts_with("cv_")) %>%
	mutate(cv_sobs = cv_sobs_2/cv_sobs_1,
			cv_shannon = cv_shannon_2/cv_shannon_1
		) %>%
	select(dataset, method, cv_sobs, cv_shannon)

@@

**The impact of removing rare ASVs on the information represented between samples.** Considering the loss of richness, diversity, and information when a community has its rarest ASVs removed, it seemed likely that the relationship between communities would also be altered. To assess the impact of removing rare ASVs on measures of alpha diversity between samples I calculated the coefficients of variation (COVs, i.e. the standard deviation divided by the mean) for richness and diversity for each study at multiple abundance thresholds. The COVs for the richness of ASVs across the studies after removing singletons were between `r oxford_comma(pull(pc_cov_sobs, cv_sobs), 1)`-times larger than they were without removing singleton ASVs (`r oxford_comma(pull(pc_cov_sobs, dataset))`; Figure 3A). Similarly, the COVs for the diversity of ASVs were between `r oxford_comma(pull(pc_cov_shannon,cv_shannon), 1)`-times larger when singletons were removed than when they were not removed (`r oxford_comma(pull(pc_cov_shannon, dataset))`; Figure 3B). To assess the impact of removing rare ASVs on measures of beta diversity between samples, I calculated the COVs of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds. The COVs between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pull(pc_cov_bc, cv_diff), 1)`-times larger than when they were not removed (`r oxford_comma(pull(pc_cov_bc, dataset))`; Figure 3C). When ASVs were clustered into OTUs the difference in COVs was less than it was for the ASVs (Figure S4). These results indicate that removing rare ASVs increases the dissimilarity between samples, which could have a significant impact on the statistical power to detect differences between treatment groups.

```{r}
b_alpha <- read_tsv(here("data/process/bffect_alpha_analysis.tsv"),
		col_types = cols(.default=col_double(),
					dataset=col_character(), method=col_character(), metric=col_character())) %>%
	filter(metric != "invsimpson", dataset != "marine") %>%
	filter(prune == 2 | prune ==1) %>%
	pivot_wider(id_cols=c(dataset, metric, method), names_from=c(prune), values_from=frac_sig) %>%
	mutate(pp_drop = `1` - `2`,
		percentage_drop = 100*(1-`2` / `1`)) %>%
	arrange(percentage_drop)

@@

Even with the best sequencing approaches, PCR artifacts and sequencing errors persist. This may account for some sequences not being observed other samples. Although this lack of inter-sample coverage could have been due to treatment effects and natural variation, it is likely that some portion of the sequences that were unique to samples were artifacts or contained errors. Researchers are encouraged to use rarefaction to control for uneven sampling and the presence of spurious sequences. Previous work sequencing mock communities has shown that the number of spurious sequences increases with sequencing depth [@Kozich2013; @Schloss2011]. By rarefying data to a common number of sequences per sample, the number of spurious sequences can be controlled. As shown in the data I presented, which was rarefied to a common number of reads per sample within a dataset, when rare sequences were not removed the power to detect differences was the highest and the false discovery rate was the expected 5% (Figures 5 and S6).

@@

**Figure 1. Singletons are more common in samples with fewer seqeunces and tend to be shared with samples having more sequences.** For each of the 12 datasets, Spearman correlation coefficients were calcualted between the number of sequences in each sample and the number of singletons in the sample (A) and the fraction of its singletons that were shared with another sample (C). Those correlations that were not statisically significant had a P-value greater than 0.05. The faction of singletons shared across samples (B) were calculated for each dataset. The median value is shown with a solid circle and the 95% confidence interval is indicated by the solid line.

@@

**Figure S2. Removing rare sequences from samples alters their representation of alpha-diversity using operational taxonomic units (OTUs).** The average difference in the richness (A), Shannon diversity (B), and Kullback-Leiber divergence (C) for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.

@@

**Figure S4. Removing rare sequences from samples increases the inter-sample variation for operational taxonomic units (OTUs).** The coefficient of variation in richness (A), Shannon diversity (B), and Bray-Curtis distances (C) for each dataset was calculated using the null distributed samples for each dataset with varying minimum abundance thresholds.

>>>>

\vspace{10mm}

**Running title:** Removal of rare ASVs biases community analyses

\vspace{25mm}

@@

```{r results_a}

frac_lost <- read_tsv(here('data/process/sequence_loss_table_raw.tsv')) %>%
	filter(freq_removed == 1) %>%
	group_by(dataset) %>%
	summarize(median=median(fraction_lost))

@@

cor_n <- read_tsv(here('data/process/sequence_loss_table_cor.tsv')) %>%
 	filter(freq_removed == 1) %>%
	select(dataset, cor_n_estimate, cor_n_p.value) %>%
	mutate(sig = cor_n_p.value < 0.05)

@@

frac_coverage <- read_tsv(here('data/process/sequence_coverage_table_raw.tsv')) %>%
	filter(freq_removed == 1) %>%
	group_by(dataset) %>%
	summarize(median=median(coverage))

@@

coverage_n <- read_tsv(here('data/process/sequence_coverage_table_cor.tsv')) %>%
	filter(freq_removed == 1) %>%
	select(dataset, cor_n_estimate, cor_n_p.value) %>%
	mutate(sig = cor_n_p.value < 0.05)

@@

**The nature of singletons.** Removal of rare ASVs is commonly justified as a method of removing ASVs that are artifacts. If such ASVs are artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth. Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`). In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figures 1A and S2). This showed that with additional sequencing, the probability of seeing singleton ASVs in multiple samples was greater than the probability of generating an artifact. This suggests that the singleton ASVs are not as likely to be artifacts as previously thought. Furthermore, if singleton ASVs were artifacts, then one would not expect to find them in other samples from the same dataset. In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences. At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset (Figure 1B). Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable. The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively ((Figures 1C and S3)). The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples. Among the three datasets without a significant correlation (Spearman correlation, P>0.05), the marine and soil datasets had the fewest samples in our collection and the stream dataset already had a high level of coverage regardless of the number of sequences. Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes more non-artifact ASVs than those that are artifacts.

```{r  results_b}
ointra <- read_tsv(here("data/process/ointra_analysis.tsv"),
		col_types = cols(.default=col_double(),
					dataset=col_character(), method=col_character(), group=col_character())
	) %>%
	group_by(dataset, method, min_class) %>%
	summarize(mean_s_frac = 100*(1-mean(s_yes / s_no)),
		mean_h_frac = 100*(1-mean(h_yes / h_no))) %>%
	ungroup()

@@

**The impact of removing rare ASVs on the information represented in each sample.** Removing rare ASVs will reduce the richness of ASVs (i.e. the number of ASVs per sample) and increase the relative abundance of the remaining ASVs. To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample. The richness of ASVs in each sample decreased by between `r oxford_comma(pull(asv_s_two, mean_s_frac), 1)`% (`r oxford_comma(pull(asv_s_two, dataset))`) when removing those ASVs that only appeared once and by between `r oxford_comma(pull(asv_s_eleven, mean_s_frac), 1)`% (`r oxford_comma(pull(asv_s_eleven, dataset))`) when removing those that appeared ten or fewer times from each sample (Figure 2A). Similarly, the Shannon diversity decreased by between `r oxford_comma(pull(asv_h_two, mean_h_frac), 1)`% (`r oxford_comma(pull(asv_h_two, dataset))`) when removing ASVs that only appeared once and by between `r oxford_comma(pull(asv_h_eleven, mean_h_frac), 1)`% (`r oxford_comma(pull(asv_h_eleven, dataset))`) when removing ASVs that appeared ten or fewer times from each sample (Figure 2B). Next, I assigned the ASVs to OTUs, which were defined as a group of ASVs that were more than 97% similar to each other to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies. Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the richness of OTUs and the diversity decreased relative to the full community (Figure S4AB). In contrast to the richness and diversity measurements, the Kullback–Leibler divergence compares the relative abundance of specific ASVs or OTUs between representations of the community. I calculated the Kullback–Leibler divergence between the full communities and those where rare ASVs were removed. As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 2C and Figure S4C). The relative loss of information was generally smaller for OTUs than than it was for ASVs. Removing rare ASVs, regardless of abundance threshold, had profound impacts on the representation of the communities.

**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing, without replacement, the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed. This effectively made every community in a study a statistical sample of the study-wide composite community distribution. For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples. Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study. The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S5). The null distribution data were used in the remainder of the study to minimize the risk of bias.

```{r results_c}
r_cov_alpha <- read_tsv(here("data/process/ralpha_analysis.tsv"),
		col_types = cols(.default=col_double(),
					dataset=col_character(), method=col_character())
	) %>% select(dataset, method, prune, starts_with("cv_")) %>%
	filter(prune == 2 | prune ==1) %>%
	pivot_wider(id_cols=c(dataset, method), names_from=prune, values_from=starts_with("cv_")) %>%
	mutate(cv_sobs = cv_sobs_2/cv_sobs_1,
			cv_shannon = cv_shannon_2/cv_shannon_1
		) %>%
	select(dataset, method, cv_sobs, cv_shannon)

@@

**The impact of removing rare ASVs on the information represented between samples.** Considering the loss of richness, diversity, and information when a community has its rarest ASVs removed, it seemed likely that the relationship between communities would also be altered. To assess the impact of removing rare ASVs on measures of alpha diversity between samples I calculated the coefficients of variation (COVs, i.e. the standard deviation divided by the mean) for richness and diversity for each study at multiple abundance thresholds. The COVs for the richness of ASVs across the studies after removing singletons were between `r oxford_comma(pull(pc_cov_sobs, cv_sobs), 1)`-times larger than they were without removing singleton ASVs (`r oxford_comma(pull(pc_cov_sobs, dataset))`; Figure 3A). Similarly, the COVs for the diversity of ASVs were between `r oxford_comma(pull(pc_cov_shannon,cv_shannon), 1)`-times larger when singletons were removed than when they were not removed (`r oxford_comma(pull(pc_cov_shannon, dataset))`; Figure 3B). To assess the impact of removing rare ASVs on measures of beta diversity between samples, I calculated the COVs of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds. The COVs between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pull(pc_cov_bc, cv_diff), 1)`-times larger than when they were not removed (`r oxford_comma(pull(pc_cov_bc, dataset))`; Figure 3C). When ASVs were clustered into OTUs the difference in COVs was less than it was for the ASVs (Figure S6). These results indicate that removing rare ASVs increases the dissimilarity between samples, which could have a significant impact on the statistical power to detect differences between treatment groups.

```{r results_d}
b_alpha <- read_tsv(here("data/process/bffect_alpha_analysis.tsv"),
		col_types = cols(.default=col_double(),
					dataset=col_character(), method=col_character(), metric=col_character())) %>%
	filter(metric != "invsimpson", dataset != "marine") %>%
	filter(prune == 2 | prune ==1) %>%
	pivot_wider(id_cols=c(dataset, metric, method), names_from=c(prune), values_from=frac_sig) %>%
	mutate(pp_drop = `1` - `2`,
		percentage_drop = 100*(1-`2` / `1`)) %>%
	arrange(percentage_drop)

@@

Even with the best sequencing approaches, PCR artifacts and sequencing errors persist. This may account for some sequences not being observed other samples. Although this lack of inter-sample coverage could have been due to treatment effects and natural variation, it is likely that some portion of the sequences that were unique to samples were artifacts or contained errors. Researchers are encouraged to use rarefaction to control for uneven sampling and the presence of spurious sequences. Previous work sequencing mock communities has shown that the number of spurious sequences increases with sequencing depth [@Kozich2013; @Schloss2011]. By rarefying data to a common number of sequences per sample, the number of spurious sequences can be controlled. As shown in the data I presented, which was rarefied to a common number of reads per sample within a dataset, when rare sequences were not removed the power to detect differences was the highest and the false discovery rate was the expected 5% (Figures 5 and S8).

@@

**Figure 1. Singletons are more common in samples with fewer seqeunces and tend to be shared with samples having more sequences.** For each of the 12 datasets, Spearman correlation coefficients were calcualted between the number of sequences in each sample and the number of singletons in the sample (A) and the fraction of its singletons that were shared with another sample (C). Those correlations that were not statisically significant had a P-value greater than 0.05. The faction of singletons shared across samples (B) were calculated for each dataset. The median value is shown with a solid circle and the 95% confidence interval is indicated by the solid line. The raw data for A and C are presented in Figures S2 and S3, respectively.

@@

**Figure S2. Singletons are more common in samples with fewer seqeunces.** Spearman correlation coefficients ($\rho$) were calcualted between the total number of sequences and the number of singletons in the sample. The Spearman correlation coefficients for the Marine and Sediment datasets were not significantly different from 0.00 (P>0.05). The blue line indicates the linear trend in the data. These data are summarized in Figure 1A.

@@

**Figure S4. Removing rare sequences from samples alters their representation of alpha-diversity using operational taxonomic units (OTUs).** The average difference in the richness (A), Shannon diversity (B), and Kullback-Leiber divergence (C) for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
